Tropical Mathematics for the Lambda Calculus

### 2023-01-15 section2.tex
commit

A natural way to account for these situations is to consider denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior. 
A fundamental insight coming from this works is that \emph{affine} programs, that is, programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive 
map \emph{symmetric monoidal closed}, and provides a model of the \emph{affine} simply typed $\lambda$-calculus.

As observed in \cite{Gabo2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications, by considering \emph{Lipschitz-continuous} functions.
For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, can be interpreted as a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$. 
By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $r\cdot X$ (i.e.~with distance $d_{r}(x,y)=r\cdot d(x,y)$) and $Y$, the program $M$ above 
can thus be interpreted as a non-expansive map from $2\cdot(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.

These observations have led to the study of \emph{bounded} typed $\lambda$-calculi (like the system $\mathsf{Fuzz}$ \cite{Reed_2010}),
which have been applied in the study of higher-order differential privacy \cite{}. In these calculi, which are inspired from Girard's \emph{bounded linear logic} \cite{}, function types are of the form $!_{r}X \multimap Y$, and programs are interpreted as $r$-Lipschitz maps. In categorical terms, the re-scaling operation $!_{r}(-)$ can be characterized as a  \emph{graded linear exponential comonad} \cite{}.

Then, what about the good old, ``unbounded'', simply typed $\lambda$-calculus? Actually, by using unbounded duplications, one might lose the Lipschitz property. For instance, while the functions $M_{k}=\lambda x. k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors. 
More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of the simply typed $\lambda$-calculus (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{}).
Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~ over any compact set). However, constructing cartesian closed categories of locally Lipschitz maps is challenging. Yet, some examples exist in the literature (see \cite{EhrhardPCOHdiff, PistoneLICS2021}), and a new example will be exhibited in this paper.

% - Metric models: functions as Lipschitz maps, differential privacy.
%
%- Problem with unrestricted duplications
%
%- Bounded Linear Logic/lambda-calculus, why Met is only a model of BLL, and why it is difficult (but not impossible!) to construct models of full lambda

@@

The key ingredient of these approaches is a notion of \emph{linear application}, which in the syntax of the differential $\lambda$-calculus is written $\Der[M]\cdot N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once. More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%

The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:
\begin{align}
MN  =  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
\end{align}
Now, given our previous discussion of bounded duplications and Lipschitz-continuity, a natural question would be whether the differential approach to the $\lambda$-calculus may lead to decompose an unbounded $\lambda$-term 
into a sum of Lipschitz functions, thus bridging the metric and differential approach.  
While suggestive, this idea seems to contradict the fact that, in the canonical semantics of the differential $\lambda$-calculus, the relational semantics \cite{}, the bounded terms appearing in the Taylor expansion correspond to \emph{polynomials}, and thus to non-Lipschitz functions.

More generally, the relational semantics interprets unbounded programs as analytic functions, that is, as functions admitting a representation as power series. For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
\begin{align}
F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
\end{align}
By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx. \sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
\begin{align}
F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
\end{align}
 In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
 $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.

Going back to our previous question, we can now reformulate it as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis? The goal of this paper is to show that this is indeed possible and actually quite natural, by appealing to tropical mathematics.

\subsection{Tropical mathematics: a possible synthesis}
The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', the computation of many algebraic and geometric objects becomes combinatorial and simple.

For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
\begin{align}
\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
\end{align}
Notably, the \emph{tropical roots} (whose definition is recalled in Section \ref{section3}) of $\trop p(\alpha)$ can be found through a rather simple (indeed polytime \cite{}) algorithm, and can be used to \emph{approximate} the actual roots of $p(x)$ \cite{}. 
More generally, the tropicalization of a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ yields a \emph{tropical Laurent series} \cite{} 
\begin{align}
\trop f(\alpha)= \inf_{n}\left\{n\alpha+ \widehat f_{n}\right\}
\end{align}
a class of functions that we will study in detail in Section \ref{section4}.

Since polynomials correspond to linear (hence Lipschitz) functions in tropical mathematics, the reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis. 
In Section \ref{section5} we will suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.  
Finally, in Section \ref{section6} we will show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between  
\emph{generalized metric spaces} (in the sense of Lawvere \cite{}) and modules over the tropical semi-ring.

%- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)

>>>>

A natural way to account for these situations is to consider denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which measures differences of behavior. 
A fundamental insight coming from this line of works is that \emph{affine} programs, that is, programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive provides a model of the \emph{affine} simply typed $\lambda$-calculus, being both a \emph{symmetric monoidal closed} category (i.e.~a model of \emph{linear} $\STLC$) and a cartesian category in which cartesian and monoidal unit coincide.

As observed in \cite{Gabo2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications. The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz-continuous} map.
For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$. 
By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $r\cdot X$ (i.e.~with distance $d_{r}(x,y)=r\cdot d(x,y)$) and $Y$, the program $M$ above 
can thus be interpreted as a non-expansive map from $2\cdot(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.

These observations have led to the study of \emph{bounded} typed $\lambda$-calculi (like the system $\mathsf{Fuzz}$ \cite{Reed_2010}), which have been applied in the study of higher-order differential privacy \cite{}. The types of such systems are defined \cite{} by combining linear constructors with a \emph{graded exponential comonad} $!_{r}(-)$ \cite{}.
In the following sections we will sometimes make reference to a basic type system, that we call $\BSTLC$, for bounded higher-order programs, with types defined as below
\begin{align}
A::= o \  \mid \  (!_{\leq n}A) \multimap A  
\end{align}
where $n$ indicates a natural number. Intuitively, $!_{\leq n}A\multimap B$ is the type of functions from $A$ to $B$ that may use their input at most $n$ times. More details about $\BSTLC$ are provided in the Appendix.

Now, what about the good old, ``unbounded'', simply typed $\lambda$-calculus? Actually, by using unbounded duplications, one might lose the Lipschitz property. For instance, while the functions $M_{k}=\lambda x. k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the function $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors. 
More abstractly, it is well-known that the category $\Met$ is \emph{not} cartesian closed, so it is not a model of $\STLC$ (yet, $\Met$ contains several cartesian closed \emph{sub-}categories, see e.g.~\cite{}).
Still, one might observe that the program $M$ above is actually Lipschitz-continuous, if not globally, at least \emph{locally} (i.e.~ over any compact set). However, constructing cartesian closed categories of locally Lipschitz maps is challenging. Yet, some examples exist in the literature (see \cite{EhrhardPCOHdiff, PistoneLICS2021}), and a new example will be exhibited in this paper.

@@

A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} (and differential linear logic) and its categorical models. 
The key ingredient is that a \emph{differential operator} $\Der$ is added to the usual syntax of the $\lambda$-calculus, with the intuition that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, which still has type $A\to B$, corresponds to a \emph{linear application} of $N$ to $M$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ has type $A\to B$, since $M$ might still ask for other copies of an input of type $A$).

Interestingly, the categorical study of such differential operators has led to introduce a family of categories, called \emph{cartesian differential categories} $C\partial C$ \cite{}, which provide an abstract axiomatization of differentiation, in the usual mathematical sense. More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
\begin{itemize}
\item $\C C$ is left-additive, i.e.~its hom-sets are commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
\item $\C C$ is equipped with a differential operator $D:
\C C(X,Y)\to \C C(X\times X,Y)$ satisfying 7 axioms $D1,\dots, D7$, which translate usual properties of the differential (e.g.~the linearity of $D$ in one of its two variables, the chain rule,  Schwarz's theorem etc.).
\end{itemize}

Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with the differential operator $\Der[-]\cdot (-)$, together with a notion of \emph{linear substitution} (see \cite{} for details).
The models of $\STDLC$ are the 
cartesian \emph{closed} differential categories ($CC\partial C$), 
which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential are both well-behaved with respect to the closed structure (see \cite{}).

Another suggestive similarity between program derivatives and 
actual derivatives is provided by the \emph{Taylor expansion}:
in $\STDLC$ one can prove that any application $MN$ can be expanded as a sum of \emph{linear} applications
$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$, as expressed below:
\begin{align}\label{eq:taylor}
MN  =  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
\end{align}
In other words, terms with unbounded duplications can be seen as some sort of ``limits'' of bounded, but arbitrarily large, duplications.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:

At this point, given our previous discussion of bounded duplications and Lipschitz-continuity, a natural question would be whether the Taylor formula
could be used to decompose an unbounded $\lambda$-term 
into a sum of Lipschitz maps, thus bridging the metric and differential approach.  
While suggestive, an obstacle to realize this idea is that,
in general, a $C\partial C$ needs not satisfy the Taylor formula
\eqref{eq:taylor}, and in the \emph{relational semantics}, 
a well-studied and sort of canonical ``Taylor'' semantics for $\STDLC$, the bounded terms appearing in the Taylor expansion correspond to \emph{polynomials}, and thus to non-Lipschitz functions.

Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis? The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to a surprising 
bridge between the metric and differential study to higher-order programs.
%
%More generally, the relational semantics interprets unbounded programs as \emph{analytic functions}, that is, as functions admitting a representation as power series. For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
%\begin{align}
%F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
%\end{align}
%By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx. \sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
%\begin{align}
%F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
%\end{align}
% In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
% $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.

\subsection{Outline of the paper}

In the next section we recall some basic ideas from tropical mathematics, and its connection with the study of the Lawvere quantale.
%Since polynomials correspond to piecewise linear (hence Lipschitz) functions in tropical mathematics,
The reconstruction of the relational semantics over the tropical semi-ring, presented in Section \ref{section3} and Section \ref{section4}, will provide a metric semantics of the differential $\lambda$-calculus, bridging sensitivity and resource analysis. 
In Section \ref{section5} we suggest potential applications of this approach, relating well-studied applications of program metrics, resource analysis with current uses of tropical mathematics in computer science.  
Finally, in Section \ref{section6} we show that the connection between the metric and differential analysis of higher-order programs extends well beyond the relational semantics, through a more abstract correspondence between  
\emph{generalized metric spaces} (in the sense of Lawvere \cite{}) and modules over the tropical semi-ring.

\section{Tropical mathematics: a possible synthesis}

At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{}, which is at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{}.
Let us recall that a quantale is a complete lattice endowed with a continuous monoid action. In our case we have $\Lawv=([0,\infty], \geq,+)$. Observe that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$. 
In particular, as discussed in Section \ref{section3}, tropical linear algebra coincides with the study of matrices with values taken over $\Lawv$.
At the same time, a (pseudo-)metric is a function $d:X\times X\to \Lawv$ satisfying usual axioms, e.g.~the triangular law (indeed, such $d$ can be seen as a $\Lawv$-\emph{enriched category}, see Section \ref{section6}).

A \emph{tropical polynomial} is a piecewise linear function $\varphi:\Lawv\to \Lawv$ of the form 
\begin{align}\label{eq:polytrop}
\varphi(\alpha)= \min_{i=1,\dots,k}\left\{ i\alpha + c_{i}\right\}
\end{align}
with coefficients $c_{i}$ taken from $\Lawv$. A value $\alpha\in \Lawv$ is a \emph{root} of the polynomial $P$ when
the minimum at $\varphi(\alpha)$ is attained at least twice (equivalently, when 
 $\varphi$ is not differentiable at $\alpha$). In other words, the tropical roots of $\varphi$ coincide with the points where the slope of $\varphi$ changes.

Intuitively, tropical polynomials look much like standard polynomials, although with ``$+$'' replaced by ``$\min$'', and ``$\times$'' replaced by ``$+$''. 
In fact, this intuition can be made precise as follows: for any positive real $t$, the tropical polynomials are in one to one correspondence with the functions $f:[0,1]\to [0,1]$ which can be written as a \emph{parameterized} polynomial 
$f_{t}(x)= \sum_{i=1}^{n}t^{c_{i}}x^{n}$, with the $c_{i}\in [0,\infty]$. 
%Hence, for any polynomial $p(x)= 
%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
%\begin{align}
%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
%\end{align}

More generally, for any function $f:[0,1]\to [0,1]$ which can be written as a (parameterized) \emph{power series} of the form $f(x)= \sum_{n}t^{c_{n}}x^{n}$ (such functions arise naturally from the interpretation of probabilistic programs, as we'll recall in Section \ref{section5}), we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$
as the function 
\begin{align}
\trop f(\alpha) =\inf_{n}\left\{ i\alpha+c_{i}\right\}
\end{align}
Such functions, called \emph{tropical Laurent series} \cite{}, will be discussed in more detail in Section \ref{section5}.

The functions $\varphi$ and $f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting
$\varphi_{t}: [0,\infty]\to [0,\infty]$ be defined by 
$\varphi_{t}(\alpha)= \phi_{t}\circ f \circ \psi_{t}$, one has that 
$\varphi= \lim_{t\to 0}\varphi_{t}$. 
Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= \log_{t}(t^{-\alpha}+t^{-\beta})$ and 
$\alpha \prodt{t}\beta:= \phi_{t}( \psi_{t}(\alpha)\psi_{t}(\beta))=
\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$ (this remark is an example of the so-called \emph{Maslov dequantization}, relating the semi-ring $\BB R$ with the tropical semiring $((-\infty, \infty], \min, +)$, see e.g.~\cite{}).

Since many algebraic and geometric properties of tropical maps are often simpler and more combinatorial than the corresponding  properties of non-tropical functions, a typical application of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.

%
%
%or more generally as a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ with coefficient $\widehat f_{n}\in [0,1]$, we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$ as the function 
%\begin{align}
%\trop f(\alpha)= \inf_{n}\left\{ 
%\end{align}
%
%This correspondence can be made precise through the the so-called
% \emph{de Maslov dequantization} \cite{}.
% For each positive real $t$, any polynomial in $\BB R[x]$ can be written under the \emph{$t$-parameterized} form:
% \begin{align}
% p_{t}(x)= \sum_{i=1}^{k}t^{c_{i}}x^{i}
% \end{align}
% with the coefficients $c_{i}$ taken from $\Lawv$. 
% It is clear then that tropical polynomials and $t$-parameterized polynomials admit a one to one correspondence between their presentations.
% 
% Actually, the $\varphi$s and the $p_{t}$s can be related by passing through some intermediate functions $\varphi_{t}$ introduced by Maslov.
%For any $t>1$, the functions $\phi_{t}(x)=-\log_{t}x$ and $\psi_{t}(\alpha)=t^{-\alpha}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
%

%
%{\color{red}Lista delle cose da dire:}
%
%1) Def di quantale, come lattice e come complete idempotent semiring.
%Among the the so-called \emph{tropical semirings}, we consider the \emph{Lawvere quantale/semiring}.
%
%2) Def di $\Lawv$, the \emph{Lawvere quantale}: seen as the idempotent complete semiring, it is $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,\cdot,0)$.
%Seen as lattice it is defined by the order $\preceq$, which is the reversed order $\geq$ of the usual order $\leq$ on $\BB R_{\geq0}\cup\set{\infty}$.
%
%3) Maslov dequantisation:
%
%First, let us recall that for any non-negative real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
%
%4) Def di $\trop$ di un polinomio/serie, (\`e la stessa formula, dipende solo se gli indici sono finiti/infiniti).
%Come sta scritto sotto, giusto un po' pi\`u formale (per esempio, scriverlo come Definizione).
%
%The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', many algebraic and geometric objects becomes combinatorial and their computation simpler. 
%
%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
%\begin{align}
%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
%\end{align}
%Notably, the \emph{tropical roots} (whose definition is recalled in Section \ref{section3}) of $\trop p(\alpha)$ can be found through a rather simple (indeed polytime \cite{}) algorithm, and can be used to \emph{approximate} the actual roots of $p(x)$ \cite{}. 
%More generally, the tropicalization of a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ yields a \emph{tropical Laurent series} \cite{} 
%\begin{align}
%\trop f(\alpha)= \inf_{n}\left\{n\alpha+ \widehat f_{n}\right\}
%\end{align}
%a class of functions that we will study in detail in Section \ref{section4}.
%
%%- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)
%
%
%
